---
title: "2: Physical Properties of Lakes"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the concepts of lake stratification and mixing by analyzing monitoring data
2. Apply data analytics skills to applied questions about physical properties of lakes
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

What are the physical properties of lakes?
temperature
depth
light transmission

Chemical properties:
ph
turbidity


## Session Set Up
```{r, message = FALSE}
# Check working directory (should be project file location)
getwd()

#install.packages("tidyverse")
#install.packages("lubridate")
install.packages("rLakeAnalyzer")

# load packages
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)

# Load data - North Temperate Lakes Long Term Ecological Research Station
NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

# set ggplot theme
theme_set(theme_classic())
```

## Data Wrangling and Exploration

### Investigate structure of dataset

Common steps/checks for data wrangling: 

* Date formatting
* Addressing NAs
* Subsetting and filtering 

```{r}
# Is the date column perceived as a date by R?
class(NTLdata$sampledate)
#its a character
NTLdata$sampledate <- as.Date(NTLdata$sampledate, format = "%m/%d/%y")
#^this function as.date changes from character to date. lower dase m/d/y is two digit, upper case is 4 digit
class(NTLdata$sampledate)
# What does this column look like now? now its a date

# Remove rows that contain NAs in temperature column
dim(NTLdata)
# gives the dimensions of the data, obs (rows) and variables (columns)
NTLdata <- NTLdata %>%
  drop_na(temperature_C) #here you dont need to use $ to indicate column
NLTdata <- drop_na(NLTdata, temperature_C) # can use drop_na without the piping
dim(NTLdata)  

# How many observations are there for each lake?
summary(NTLdata$lakename)
summary(as.factor(NTLdata$lakename)) #a count of how many time each lake name appears in the lakename column. becasue we removed missed temp data, this tells up how many temp readings we have for each lake

# Let's choose the two lakes with the most data
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake")) #filter takes whats included. %in% = include, c = what to include

# What is another way to use the filter command to get to the same result?
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") # | means "or"

# Make two data frames, one for each lake
Pauldata <- NTLdata %>%
  filter(lakename == "Paul Lake")
Peterdata <- NTLdata %>%
  filter(lakename == "Peter Lake")

# How long did the monitoring last? 
min(Pauldata$sampledate)
max(Pauldata$sampledate)
min(Peterdata$sampledate)
max(Peterdata$sampledate)

# Which depths are sampled in each lake?
unique(Pauldata$depth) #doesnt put them in exact order, unique pulls each value once
unique(Peterdata$depth) 
summary(Peterdata$depth) # doesnt show show up the increments at which they sample
# Why didn't we use the "summary" function here?

# QA the one data point at 20 m depth. 
Pauldata <- Pauldata %>%
  filter(depth < 20)

```

### Exploratory data visualization

Let's make a plot of temperatures by depth. There are a lot of points, so adding a 50 % transparency to the points helps us see where points are densely clustered together.

Let's also try using the traditional limnological graph type, with depth on the y axis in reverse, to simulate a cross section of a lake. When and where do we usually observe high and low temperatures?
```{r}
ggplot(Pauldata, aes(x = depth, y = temperature_C)) + #aes asthetics = what do you want to inlcude
  geom_point(alpha = 0.5) + #alpha = tranparency
  labs(y = expression("Temperature "(degree*C)), x = "Depth (m)")

ggplot(Pauldata, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY")
#this gives us a more cross section view of the lake. this also add the color which shows day of year
```

How do temperatures at the surface compare to temperatures at the mid-depths and at the bottom?
-temp decreases with depth
-temp varies more at the surface than deeper depths
-temp is the same at all depths at coldest temp = thermally mixed. the lake has very little resistence to mixing
-stratified
  epilimnion=upper water column
  hypolimnion= lower water column
  in warmer months density of water due to temp resists mixing
-what tends to happen across a year:
  winter - frozen on top then increases to 4 C water is densest - it stableizes here. 
  spring - same temo through out, spring and fall looks the same
  summer- looks like second ggplot
  fall- water colimn mixes, you get consistent temp throughout depth
  this is a Dimictic lake
-monomictic lake:
fall, winter, spring, same temo through out
summer, looks like second ggplot, mixed temp throughout

-polymictic lake:
mixes several times per year, not seasonal. mixes with weather patterns, like wind.
shallower lakes are more likely polymictic, less thermal mass

-amictic lake:
lakes that never mix, always statefied. very sheltered. 

### Graphing seasonal water profiles

Lakes in the North Temperate Lakes LTER are dimictic, meaning they mix fully twice per year. When the lakes are not mixed (stratified), the top layer (epilimnion) and the bottom layer (hypolimnion) of the lake are different temperatures.

For background information on lake stratification, see this [stratification summary article from the North American Lake Management Society](https://www.nalms.org/secchidipin/monitoring-methods/temperature-and-oxygen/).


In 1993, Paul Lake was sampled from May 20 through November 2. Let's look at temperature profiles of the lake across the year and also explore how this impacts dissolved oxygen.

```{r}
ggplot(subset(Pauldata, year4 == 1993), 
       aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY") 

# Now, make a plot as above but with dissolved oxygen saturation instead of temperature.
ggplot(subset(Pauldata, year4== 1993),
       aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("DO (mg/L)", y = "Depth (m)", color = "DOY"))

```
plot analyzation:
-higher DO in water colomn during the day
-bulge of algae right below thermocline around 3m
-phytoplankton need light AND mutrients, low nutrients in upper thermocline  - lots of competition. phytoplankton hang out where they are still getting enough light but where there is more nutrients

Why does mixing occur in the spring and fall? What are the mechanisms that make this possible?
<add notes here>
density difference gets smaller which allows mixing




### Calculating thermoclines
thermocline - where temp is changin - where you see greater than 1 degree of temp over less than 1 m depth

[rLakeAnalyzer documentation](https://cran.r-project.org/web/packages/rLakeAnalyzer/rLakeAnalyzer.pdf)

```{r}
Pauldata_thermo <- Pauldata %>%
  group_by(year4, daynum, sampledate) %>%
  # calculate thermoclines based on temperature profiles. 
  # seasonal = FALSE calculates the thermocline as the maximum density gradient
  # seasonal = TRUE calculates the thermocline as the deepest density gradient
  summarise(thermocline = thermo.depth(wtr = temperature_C, depths = depth, seasonal = FALSE)) %>%
  #takes temperatiure profile and depths associated with those temps
  # remove all thermoclines within 1 m of the surface. these can represent transient stratification.
  filter(thermocline > 1)

ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline)) +
  geom_point() 
  
```

How does thermocline depth change over the course of the year? When do we observe the most variability in thermocline depth?
-thermocline deepens as year goes on
-day 0 = first stratification

### Climate change

Increases in global temperature are predicted to cause several changes to lake thermal conditions, including: 

* Increases in surface temperature
* Increases in surface minimum temperature
* Increases in extent of stratification
* Increases in length of stratification
* Decreases in ice cover

Several studies have addressed this topic, using long-term and spatially diverse datasets: 

https://link.springer.com/article/10.1007/s10584-015-1326-1?sa_campaign=email/event/articleAuthor/onlineFirst&error=cookies_not_supported&error=cookies_not_supported&code=2b415e25-de4c-452f-bd02-2cceae08b7a3&code=e63aabb9-76d3-4e49-b36c-e591007a9e9c

http://hpkx.cnjournals.com/uploadfile/news_images/hpkx/2020-07-15/10.1038-s43017-020-0067-5.pdf

https://agupubs.onlinelibrary.wiley.com/doi/pdf/10.1002/2015GL064097

https://link.springer.com/article/10.1007/s10584-019-02465-y

Let's explore how these changes might manifest in Paul Lake. 
```{r}
ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline, color = year4)) +
  geom_point() +
  scale_color_viridis_c()
# exercise: put the legend on top and relabel the aesthetics

#how might we expect the thermocline to change over the year?
  #-deepening earlier in the season
  #need to search trhough the noise of the data set

ggplot(subset(Pauldata, depth == 1),
       aes(x = daynum, y = temperature_C, color = as.factor(year4))) +
  geom_point(alpha = 0.5) +
  geom_line() + 
  scale_color_viridis_d()
# exercise: relabel the aesthetics



```

## Closing Discussion

What are the main concepts you learned about the physical properties of lakes today? What was the evidence for these concepts in the dataset?
