---
title: "5: Physical Properties of Rivers"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze stormflow hydrographs in the context of baseflow and stormflow
2. Conduct baseflow separation on discharge data
3. Diagnose and communicate hydrologic changes in the context of changing baseflows

## Opening Discussion

What might cause streams to have higher peak flow following a precipitation or snowmelt event? 
-slope effects time of concentration = discharge over period of time
-impervious surface
-magnitude of storm


## Session Set Up

```{r, message = FALSE}
getwd()

# install.packages("EcoHydRology")

library(EcoHydRology) #order in which you load your packages, determines hierarchy of functions
library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(lfstat)

theme_set(theme_classic())
```


## Hydrograph Analysis and Baseflow Separation for the Eno River

Let's import discharge data for the Eno River near Durham for all full water years. This should look familiar based on previous lessons.

```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "1963-10-01", 
                     endDate = "2021-09-30") #complete water years

names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

attr(EnoDischarge, "variableInfo")
attr(EnoDischarge, "siteInfo")
  
ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")
```

Notice that the Eno River is very responsive to precipitation events, with frequent discharge peaks throughout the period of record. How much of the flow of the Eno River is attributed to stormflow pathways (also called quickflow) vs. baseflow?

### Calculating baseflow and stormflow

**Stormflow**(also called quick flow) is attributed to overland flow and shallow subsurface flow that deliver water quickly to a stream or river of interest. **Baseflow** is attributed to slower flow pathways, including but not limited to groundwater. Sometimes snowmelt is considered base flow.

The `lfstat` package includes a calculation for baseflow (function: `baseflow`). The `EcoHydRology` package also includes a `BaseflowSeparation` function. Both of these packages have optional specifications that can be tinkered with by the user based on the specific context of the stream of interest. See the help files for these functions for more info.

```{r}
# calculate baseflow with the lfstat package, and add Year and Water Year
EnoDischarge <- EnoDischarge %>%
  mutate(Baseflow_lfstat = baseflow(Discharge), 
         Stormflow_lfstat = Discharge - Baseflow_lfstat,
         Year = year(Date), 
         WaterYear = water_year(Date))#interpolates base flow, so we will not see base flow for the first and last 12 
EnoDischarge$WaterYear <- as.numeric(as.character(EnoDischarge$WaterYear)) #change from factor to numeric via character

# plot baseflow and total flow
ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
  geom_line(color = "gray") +
  geom_line(aes(y = Baseflow_lfstat)) + #need to specify a new y aesthetic
  scale_y_log10() +
  labs(x = "Year", y = "Discharge (cfs)")
#base flow mirror discharge, which make sense, in the summer we will see base flows decrease...?

#example of how to add a geomline from another dataset altogether
#ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
  #geom_line(color = "gray") +
  #geom_line(data = otherdataset, aes(y = Baseflow_lfstat)) + #need to specify a new y aesthetic
  # scale_y_log10() +
  #labs(x = "Year", y = "Discharge (cfs)")



# calculate baseflow with the EcoHydRology package.
# Note: this function creates two columns which can be joined onto the original dataset.calcualtes stormflow and baseflow. bt = baseflow, qft= quick flow
EnoDischarge_basesep_EcoHydro <- BaseflowSeparation(EnoDischarge$Discharge) 

EnoDischarge <- cbind(EnoDischarge, EnoDischarge_basesep_EcoHydro)

# Let's compare the two approaches. How well do they match up, and is one biased a certain way?
ggplot(EnoDischarge, aes(x = Baseflow_lfstat, y = bt)) +
  geom_point(alpha = 0.1, size = 1) + #alpha=transparency
  geom_abline(slope = 1, intercept = 0, color = "darkred") #absolute line, 1:1 line. closer the points are to the line the more similar the baseflow interpolation was.

# How would we test that statistically? [add code here]
#paired t test - this compares the means of the groups, differences in coefficeints could infer
#linear model - showed not significant differnce
summary(lm(data=EnoDischarge, bt ~ Baseflow_lfstat)) #R2 valued = variation from line of best fit, 93 % of variablity is accounted for. Slope term of "Estimate" .93 is good. note that intercept is a .6, so not at 0. this all means these methods are fairly similar in their total impact. 


```

### Seasonal and event-based cycles

Let's zoom in on a few recent years to look at how baseflow and stormflow vary across seasonal cycles as well as discrete events. 

[add notes here about our exploration]

```{r}
EnoDischarge_18_20 <- EnoDischarge %>%
  filter(WaterYear %in% c(2018:2020))

ggplot(EnoDischarge_18_20, aes(x = Date, y = Discharge)) +
  geom_line(color = "gray") +
  geom_line(aes(y = Baseflow_lfstat), color = "darkred") +
  geom_line(aes(y = bt)) +
  scale_y_log10() +
  labs(x = "Year", y = "Discharge (cfs)")
#we notice: difference in base flow calcualtions, different levels of smoothing happening. You can set maximums and smoothing in the function. We can see how baseflow increases in response to storm events. 

EnoDischarge_ex<- EnoDischarge %>%
  filter(Date > "2019-07-15" & Date < "2019-09-01")

ggplot(EnoDischarge_ex, aes(x = Date, y = Discharge)) +
  geom_line(color = "gray") +
  geom_line(aes(y = Baseflow_lfstat), color = "darkred") +
  geom_line(aes(y = bt)) +
  scale_y_log10() +
  labs(x = "Year", y = "Discharge (cfs)")

```

### Annual statistics

Now, let's calculate some annual flow metrics for the Eno. We can add up daily discharge to calculate annual discharge. 

```{r}
EnoDischarge_summary <- EnoDischarge %>%
  group_by(WaterYear) %>% #groupby with summarise, runs stats on groups data
  summarise(Discharge.acft.yr = sum(Discharge)*723.968, # what is this factor? changes cubic*ft/sec to acre*ft/year
            Baseflow.acft.yr = sum(bt)*723.968,
            Stormflow.acft.yr = sum(qft)*723.968,
            Prop.bf = Baseflow.acft.yr/Discharge.acft.yr, #porportion 
            Prop.sf = Stormflow.acft.yr/Discharge.acft.yr) %>%
 mutate_if(is.numeric, round, 2) # mutate_if, if a column is numeric, r will round to 2 decimals. Not missing data in first and late year due to baseflow calc


ggplot(EnoDischarge_summary, aes(x = WaterYear, y = Discharge.acft.yr)) +
  geom_line() +
  geom_line(aes(y = Baseflow.acft.yr), lty = 2) #higher storm flow relates to higher base flow. lty = line time (dotted, solid etc.)

ggplot(EnoDischarge_summary, aes(x = WaterYear, y = Prop.bf)) +
  geom_col() #good to look at porporiton over time for annual stats


```

## Arkansas River Example: Altered baseflows

Split into two groups. Each group will acquire discharge data and perform a baseflow separation analysis for a gage on the Arkansas River in Kansas. You may choose to use the baseflow function from the `lfstat` or the `EcoHydRology` package. After you've calculated baseflow, create a graph of daily discharge by date and then add baseflow as another line. 

* Group 1: gage 07137500  in Coolidge, Kansas 
* Group 2: gage 07139500 in Dodge City, Kansas
* Both groups: Daily mean discharge
* Both groups: 1960-2005

After you've completed your task, find a partner who analyzed the other gage. Compare the patterns in discharge and baseflow between the two sites. Work together to make a diagnosis of what could be happening over time in the Arkansas River (hint: the Coolidge site is upstream of the Dodge City site). 

[Add notes here from your discussion with your partner and our discussion as a class]

This exercise is based off a case study in Zimmer et al. 2020: (Zero or not? Causes and consequences of zero-flow stream gage  readings)[https://wires.onlinelibrary.wiley.com/doi/10.1002/wat2.1436]

```{r}
ArkDischarge <- readNWISdv(siteNumbers = "07137500",
                     parameterCd = "00060", # code for discharge (ft3/s)
                     startDate = "1960-10-01", 
                     endDate = "2005-09-30") #complete water years

names(ArkDischarge)[4:5] <- c("Discharge", "Approval.Code")

ggplot(ArkDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")

```

```{r}
ArkDischarge_basesep_EcoHydro <- BaseflowSeparation(ArkDischarge$Discharge) 

ArkDischarge <- cbind(ArkDischarge, ArkDischarge_basesep_EcoHydro)
```


```{r}
ggplot(ArkDischarge, aes(x = Date, y = Discharge)) +
  geom_line(color = "gray") +
  geom_line(aes(y =  bt), color = "darkred") +
  scale_y_log10() +
  labs(x = "Year", y = "Discharge (cfs)")
```

Because of the groundwater pumping, the river became a loosing reach (loosing water to ground water through the river). the baseflow became non-prenial, it stops flowing. 