---
title: "Assignment 9: Mapping"
author: "Natalie von Turkovich"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on mapping

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/NDWEUu73LooFJPVM8 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-04-12

## Setup

1. Load the tidyverse, LAGOSNE, maps, dataRetrieval, sf, and nhdplusTools packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
library(tidyverse)
library(dataRetrieval)
library(maps)
library(sf)
#install.packages("nhdplusTools")
library(nhdplusTools)
library(LAGOSNE)

theme_set(theme_classic())
options(scipen = 5)

```

## LAGOS-NE

2. Choose five lakes in the LAGOS-NE database that are located within the same state and have chlorophyll data. Subset your data accordingly, and create two plots:

* A map of the five lakes within the state boundary, with each lake point as its own color.
* A boxplot with jittered points layered on top of chlorophyll concentrations in each lake (chlorophyll on y axis, lake on x axis), with each lake having a fill and/or color that matches the map. 

```{r}
LAGOSdata <- lagosne_load()

states <- st_as_sf(map(database = "state", plot = FALSE, fill = TRUE, col = "white"))

states.subset <- filter(states, ID== "vermont")

LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr
LAGOScombined <-
  left_join(LAGOSnutrient, LAGOSlocus) %>%
  left_join(., LAGOSstate) %>%
  filter(state == "VT") %>%
  select(lagoslakeid, gnis_name, sampledate, chla, nhd_lat, nhd_long) %>%
  drop_na(chla) %>%
  arrange(lagoslakeid)

lakesample <- sample(x = LAGOScombined$lagoslakeid, size = 5) #random sample of 5 lakes

LAGOSsample <- LAGOScombined %>%
  filter(lagoslakeid %in% lakesample)

LAGOSsample.spatial <- st_as_sf(LAGOSsample, coords = c("nhd_long", "nhd_lat"), crs = 4326)
      
#map                          
ggplot() +
  geom_sf(data = states.subset, fill = "white") +
  geom_sf(data = LAGOSsample.spatial, aes(color = as.factor(lagoslakeid))) +
  scale_color_viridis_d() +
  labs(color = "Lake ID") +
  theme(legend.position = "top")

#scatter box
ggplot(LAGOSsample, aes(x = as.factor(lagoslakeid), y = chla, fill = as.factor(lagoslakeid)))+
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  scale_y_log10() +
  scale_fill_viridis_d() +
  labs(x = "Lake ID", y = expression("Chlorophyll a ("*mu*"g/L)"), fill = "") +
  theme(legend.position = "none")
```


## NHDPlus

3. Delineate and map the watershed and flowpaths in the network upstream of the Eno River gage in Durham (USGS-02085070). Your map should include all stream orders.


```{r}
# Recall there are a lot of parameters measured at the site
EnoParams <- whatNWISdata(siteNumbers = "02085070")

# Extract latitude and longitude for the site
EnoCoords <- EnoParams %>%
  select(site_no, dec_lat_va, dec_long_va) %>%
  distinct()

# Define the gage site as the starting point
start_point <- st_sfc(st_point(c(EnoCoords$dec_long_va, EnoCoords$dec_lat_va)), crs = 4269) # NAD83, commonly used by US agencies

start_comid <- discover_nhdplus_id(start_point)
# start_point2 <- st_as_sf(data.frame(x = NeuseCoords$dec_long_va, y = NeuseCoords$dec_lat_va),
# coords = c("x", "y"), crs = 4269)

# Navigate the NLDI network
NLDI <- navigate_nldi(list(featureSource = "comid", featureID = start_comid),
                          mode = "upstreamTributaries",
                          distance_km = 1000) # Extract watershed and flowpath information

# Extract watershed and flowpath information
subset_file <- tempfile(fileext = ".gpkg")
subset <- subset_nhdplus(comids = as.integer(NLDI$UT$nhdplus_comid),
                         output_file = subset_file,
                         nhdplus_data = "download",
                         flowline_only = FALSE,
                         return_data = TRUE, overwrite = TRUE)

#Warning message:No nhdarea features found  - this is OK


# Create data frames
flowline <- subset$NHDFlowline_Network
catchment <- subset$CatchmentSP
plot_nhdplus("USGS-02085070")

max(flowline$totdasqkm)

summary(as.factor(flowline$streamorde))

gages <- get_nwis(AOI = catchment)

gages <- st_intersection(gages, catchment)
```

4. What is the upstream drainage area for this gage, and how are the flowpaths distributed with regard to stream order?

> The upstream drainage area is 367.68 square kilometers. There are 123 flowpaths of stream order 1, 45 of stream order 2, 19 of stream order 3, and 28 of stream order 4.

5. Are there any other gage sites in the upstream catchment? If so, where are they?
> Yes there are two other gages, one is located in Hillsborough and on Cole Mill Road.
